Transfer of Learned Opponent Models in Zero Sum Games

### 2020-09-04 Experiment_2/Experiment_2_modelling.Rmd
fixed Q learning bugs

## Fitting Q_learning with state space of possible last round plays 
```{r, cache=TRUE}
exp2_QL_states <- function(par,data,return_value,gamma, within_transfer = FALSE){
  # Par[1] -> beta= inverse temperature parameter in softmax choice function
  # Par[2] -> lambda = learning rate (one per game?)
  # data : participant and ai choice data.
  # return_value=c("-2loglik","likelihood_by_trial")
  # Gamma is discount factor for future rewards 
  # Returns Q-values per trial and predicts choice using softmax
  beta <- par[1]
  alpha <- par[2]

@@

  for(t in 1:nrow(data)) {
    t_game <- data[t,"game"]
    nopts <- 3

    if(data[t,"round"] == 1) {

@@

      #cat(Q_vals,"\n") 
      state_vec <- switch(as.character(t_game),rps=states_RPS,fwg = states_FWG, shootout = states_SHOOT)
      # first round is uniform prediction
      lik_hum[t] <- 1/nopts
      # Randomly select state and previous actions for first round
      curr_state <- sample(state_vec, size = 1)
      h_act <- sample(colnames(Q_vals), size =1)
      ai_act <- sample(colnames(Q_vals), size =1)
      reward <- 0

    # Q_learning: update rule (time  = t-1)
    Q_vals[curr_state, h_act] <- Q_vals[curr_state, h_act] + alpha*( reward + gamma*max(Q_vals[new_state,]) - Q_vals[curr_state, h_act] )

    # Assume human chooses action probabilistically using softmax on Q values
    probs <- exp(Q_vals[new_state,]/beta)/sum(exp(Q_vals[new_state,]/beta))
    #if (data[t,"round"] == 1) { cat(probs,"\n") }

    # Update state
    curr_state <- new_state 
    #}
  }
  if(return_value == "-2loglik") {
    ret <- -2*sum(log(lik_hum))
    if(is.infinite(ret) || is.nan(ret)) {
      return(1e+300)
    } else {
      return(ret)
    }
  }
  if(return_value == "likelihood_by_trial") return(lik_hum)

>>>>

## Fitting Q_learning with state space of possible last round plays 
# exp2_QL_states <- function(par,data,return_value,gamma, within_transfer = FALSE){
#   # Par[1] -> beta= inverse temperature parameter in softmax choice function
#   # Par[2] -> lambda = learning rate (one per game?)
#   # data : participant and ai choice data.
#   # return_value=c("-2loglik","likelihood_by_trial")
#   # Gamma is discount factor for future rewards 
#   # Returns Q-values per trial and predicts choice using softmax
#   beta <- par[1]
#   alpha <- par[2]
# 
#   #gamma <- 0.9 
# 
#   #Define matrix of state spaces for each game 
#   G1 <- expand.grid(c("R", "P", "S"),c("R", "P", "S"))
#   states_RPS <- paste0(G1$Var1,G1$Var2)
# 
#   G2 <- expand.grid(c("F", "W", "G"), c("F", "W", "G"))
#   states_FWG <- paste0(G2$Var1,G2$Var2)
# 
#   G3 <- expand.grid(c("left","center","right"), c("left","center","right"))
#   states_SHOOT <- paste0(G3$Var1,G3$Var2)
#   
#   Q_vals_RPS = matrix(-0.5,9,3)
#   dimnames(Q_vals_RPS) = list(states_RPS, c("R", "P", "S"))
# 
#   Q_vals_FWG = matrix(-0.5,9,3)
#   dimnames(Q_vals_FWG) = list(states_FWG, c("F", "W", "G"))
#   
#   Q_vals_SHOOT = matrix(-0.5,9,3)
#   dimnames(Q_vals_SHOOT) = list(states_SHOOT, c("left","center","right"))
#  
#   lik_hum <- matrix(0.0,nrow(data))
# 
#   for(t in 1:nrow(data)) {
#     t_game <- data[t,"game"]
#     nopts <- 3
#     
#     if(data[t,"round"] == 1) {
#       
#       # Within game transfer Code 
#       if (!within_transfer){
#               Q_vals <- switch(as.character(t_game),rps=Q_vals_RPS,fwg = Q_vals_FWG, shootout = Q_vals_SHOOT)
#       } else {
#         if (data[t,"stage"] == 1){
#           Q_vals <- switch(as.character(t_game),rps=Q_vals_RPS,fwg = Q_vals_FWG, shootout = Q_vals_SHOOT)
#         } else if (data[t,"stage"] == 2) {
#           Q_vals_opp1 <- Q_vals
#           Q_vals <- switch(as.character(t_game),rps=Q_vals_RPS,fwg = Q_vals_FWG, shootout = Q_vals_SHOOT)
#         } else if (data[t,"stage"] == 3) {
#           Q_vals_opp2 <- Q_vals 
#           Q_vals <- Q_vals_opp1
#         } else if (data[t,"stage"] == 4) {
#           Q_vals <- Q_vals_opp2
#         }
#       }
#    
# 
#       #cat(Q_vals,"\n") 
#       state_vec <- switch(as.character(t_game),rps=states_RPS,fwg = states_FWG, shootout = states_SHOOT)
#       # first round is uniform prediction
#       lik_hum[t] <- 1/nopts
#       # Randomly select state and previous actions for first round
#       curr_state <- sample(state_vec, size = 1)
#       h_act <- sample(colnames(Q_vals), size =1)
#       ai_act <- sample(colnames(Q_vals), size =1)
#       reward <- 0 
#       
#     } else {
#       # Get past human action and associated reward 
#       h_act <- as.character(data[t-1,"h_action"])
#       ai_act <- as.character(data[t-1,"a_action"])
#       reward <- as.numeric(data[t-1,"score"])
#     }
#     
#     
#     # cat(curr_state, " This is the previous state", "\n")
#     new_state <- paste0(h_act,ai_act)
#     # cat(new_state,"This is new state","\n")
#     
#     # Q_learning: update rule (time  = t-1)
#     Q_vals[curr_state, h_act] <- Q_vals[curr_state, h_act] + alpha*( reward + gamma*max(Q_vals[new_state,]) - Q_vals[curr_state, h_act] )
#     
#     # Assume human chooses action probabilistically using softmax on Q values
#     probs <- exp(Q_vals[new_state,]/beta)/sum(exp(Q_vals[new_state,]/beta))
#     #if (data[t,"round"] == 1) { cat(probs,"\n") }
#     
#     # Get actual human action and compute likelihood
#     h_act <- as.character(data[t,"h_action"])
#     act_index <- match(h_act, colnames(Q_vals))
#     lik_hum[t] <- probs[[act_index]]
#       
#     # Update state
#     curr_state <- new_state 
#     #}
#   }
#   if(return_value == "-2loglik") {
#     ret <- -2*sum(log(lik_hum))
#     if(is.infinite(ret) || is.nan(ret)) {
#       return(1e+300)
#     } else {
#       return(ret)
#     }
#   }
#   if(return_value == "likelihood_by_trial") return(lik_hum)
#   
# }

```
## Attempt 2 : QL with states and option to activate transfer 
```{r, cache=TRUE}
exp2_QL_states <- function(par,data,return_value,gamma, within_transfer = FALSE){
  # Par[1] -> beta= inverse temperature parameter in softmax choice function
  # Par[2] -> lambda = learning rate (one per game?)
  # data : participant and ai choice data.
  # return_value=c("-2loglik","likelihood_by_trial")
  # Gamma is discount factor for future rewards 
  # Returns Q-values per trial and predicts choice using softmax
  beta <- par[1]
  alpha <- par[2]
  nopts <- 3

@@

  for(t in 1:nrow(data)) {
    t_game <- data[t,"game"]
    if(data[t,"round"] == 1) {

@@

      #cat(Q_vals,"\n") 
      state_vec <- switch(as.character(t_game),rps=states_RPS,fwg = states_FWG, shootout = states_SHOOT)
      # first round is uniform prediction
      lik_hum[t] <- 1/nopts

    } else {
      # Get past human action and associated reward 
      h_act_prev <- as.character(data[t-1,"h_action"])
      ai_act_prev <- as.character(data[t-1,"a_action"])
      curr_state <- paste0(h_act_prev,ai_act_prev)

      h_act <- as.character(data[t,"h_action"])
      ai_act <- as.character(data[t,"a_action"])
      reward <- as.numeric(data[t,"score"])
      new_state <- paste0(h_act,ai_act)

      # Assume human chooses action probabilistically using softmax on Q values
      probs <- exp(Q_vals[curr_state,]/beta)/sum(exp(Q_vals[curr_state,]/beta))
      # Get actual human next action and compute likelihood
      act_index <- match(h_act, colnames(Q_vals))
      lik_hum[t] <- probs[[act_index]]

      # Q_learning: update rule (time = t)
      Q_vals[curr_state, h_act] <- Q_vals[curr_state, h_act] + alpha*( reward + gamma*max(Q_vals[new_state,]) - Q_vals[curr_state, h_act])
    }

  }
  if(return_value == "-2loglik") {
    ret <- -2*sum(log(lik_hum))
    if(is.infinite(ret) || is.nan(ret)) {
      return(1e+300)
    } else {
      return(ret)
    }
  }
  if(return_value == "likelihood_by_trial") return(lik_hum)

### 2020-09-04 Experiment_2/exp2_lik_by_trial.Rmd
fixed Q learning bugs

# dat <- as_tibble(dat) %>% group_by(human_id,game)
  dat <- dat %>%
   mutate(ai_action_prev = lag(ai_action,1), human_action_prev = lag(human_action,1))

@@

lik_by_trial_2 <- data.frame()

@@

  # QLS likelihoods 
  pars_QLS <- unlist(filter(exp2_optim_results, ID == id)%>%select(QL_states_alpha,QL_states_Beta))
  tdat$QLS_lik <-  exp2_QL_states(pars_QLS,tdat,"likelihood_by_trial",0.0, FALSE)

  # QLS within Transfer likelihoods 
  pars_QLS_within <- unlist(filter(exp2_optim_results, ID == id)%>%select(QLS_within_Tr_alpha,QLS_within_Tr_Beta))
  tdat$QLS_lik_within <-  exp2_QL_states(pars_QLS_within,tdat,"likelihood_by_trial",gamma = 0, TRUE)
  #                
  # 
  # Bayes model no transfer likelihoods
  theta_NT <- unlist(filter(exp2_optim_results, ID == id)%>%select(theta_no_tr))
  tdat$Bayes_NT_lik <- Bayes_model_LL(theta_NT, tdat, distinct_opponent = TRUE, generalize = "no",return = "likelihood_by_trial")

>>>>

dat <- as_tibble(dat) %>% group_by(human_id,game)
dat <- dat %>%
   mutate(ai_action_prev = lag(ai_action,1), human_action_prev = lag(human_action,1))

@@

dat <- as.data.frame(dat)   
lik_by_trial_2 <- data.frame()

@@

  # QLS likelihoods 
  pars_QLS <- unlist(filter(exp2_optim_results, ID == id)%>%select(QL_states_Beta, QL_states_alpha))
  tdat$QLS_lik <-  exp2_QL_states(pars_QLS,tdat,"likelihood_by_trial",0.0, FALSE)

  # QLS within Transfer likelihoods 
  pars_QLS_within <- unlist(filter(exp2_optim_results, ID == id)%>%select(QLS_within_Tr_Beta,QLS_within_Tr_alpha))
  tdat$QLS_lik_within <-  exp2_QL_states(pars_QLS_within,tdat,"likelihood_by_trial",gamma = 0, TRUE)
  #                
  # 
  # Bayes model no transfer likelihoods
  theta_NT <- unlist(filter(exp2_optim_results, ID == id)%>%select(theta_no_tr))
  tdat$Bayes_NT_lik <- Bayes_model_LL(theta_NT, tdat, distinct_opponent = TRUE, generalize = "no",return = "likelihood_by_trial")

### 2020-09-04 Report/draft_report_v1.tex
fixed Q learning bugs

Next we compared the performance of players whose actions are consistent with each of our hypothesized models. Figure~\ref{fig:exp1-cumScores} shows the average cumulative performance of players across games, for participants grouped by which model best fits their behavior in experiment 1. We can see that participants whose actions are most consistent with learning a ToM opponent model in a Bayesian way had the best overall performance (both with and witout transfer), followed by Q-learning conditional on last round play. EWA, QL and random players had, understandably the lowest performance.

>>>>

Next we compared the performance of players whose actions are best fit by each of our hypothesized models. Figure~\ref{fig:exp1-cumScores} shows the average cumulative performance of players across games, for participants grouped by which model best fits their behavior in experiment 1. We can see that participants whose actions are most consistent with learning a ToM opponent model in a Bayesian way had the best overall performance (witout transfer), followed by Q-learning conditional on last round play. EWA, QL and random players had, understandably the lowest performance.
